# A woman-friendly welfare state?

Regarding overall female employment rates, there are substantial differences between welfare state regimes. In the 1980s, an M-shaped labor-force pattern was prevalent in the Central European, Anglo-liberal, and East Asian countries; a sharply peaked pattern dominated in Southern Europe, while the Nordic pattern followed a reverse U-shape. Traditionally, having children in Southern Europe led to giving up employment, which was reflected in the peak employment figures for females in the 20–24 age bracket, after which, in older age groups, employment rates sharply diminished. The previous M-shape for the Anglo-liberal regime indicated that women with family responsibilities were outside the labor force, but once their children had grown up, mothers resumed working.

Something has changed since the 1980s. All other countries have “gone Nordic” and their labor-market participation profiles are approaching the Nordic ones, i.e., there is an ongoing process of convergence. Countries are becoming more similar. Central European and former socialist countries display high participation rates but the participation profile is much narrower; that is, women enter later and leave the labor market earlier than in Scandinavia. In Anglo-Saxon, “liberal” countries, levels are lower but the profile is wider (with the exception of Ireland). Whereas in South Europe, the participation profile still is peaked and narrow.

In Figure 6, female labor-force participation is inspected from a different angle. We concentrate on those women whose labor-force status is either “taking care of children or other relatives” or “inactive.” Data are derived from the EU Statistics of Income and Living Conditions (Eurostat, 2011). The figure displays female care responsibilities along the whole lifespan from 20 to 65 years of age. As can be seen, whereas in Denmark, Norway, and Sweden inactive spells are negligible and short, in the Czech Republic, Estonia, Finland, Hungary, Poland, and the United Kingdom the most intensive periods of care are in the 25–45 age bracket. Likewise for Iceland, but in Iceland there is a tendency that care-related absences also increase in older age brackets. In the Czech Republic and the United Kingdom the absence spells tend to be somewhat longer. Southern Europe and the Benelux countries – particularly Luxembourg – display a distinct pattern, where family-related care tasks increase substantially with age and the absence spells simultaneously become much longer.

```{r figuuri6}
if (!file.exists("./local_data/eusilc_trimmed.Rdata")){




  if (!file.exists("C:/btsync/data/Eurostat/EU-SILC/Longitudinal 2011/longi2011rev1/longi2011rev1_personal_all.RData")){

    library(r.eusilc)
    fldr <- "C:/btsync/data/Eurostat/EU-SILC/Longitudinal 2011/longi2011rev1/" # local C
    merge_eusilc(path.personal.register  = paste0(fldr,"/UDB_l11R_ver 2011-2 from 01-03-2014.csv"),
                 path.personal.data      = paste0(fldr,"/UDB_l11P_ver 2011-2 from 01-03-2014.csv"),
                 output.path="C:/btsync/data/Eurostat/EU-SILC/Longitudinal 2011/longi2011rev1/",
                 output.filename = "longi2011rev1_personal_all",
                 return.object = FALSE,
                 level="personal",
                 format="RData",
                 subset.vars.per.reg="all",
                 subset.vars.per.data="all",
                 subset.countries="all")

  } else load("C:/btsync/data/Eurostat/EU-SILC/Longitudinal 2011/longi2011rev1/longi2011rev1_personal_all.RData")

  dat <- merged %>%
    filter(!is.na(PL031)) %>%
    group_by(RB020,RB030) %>%
    mutate(max = max(RX010, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(age_cat = cut_interval(x = max, length = 5)) %>%
    mutate(sex = ifelse(RB090 == 1, "male",
                        ifelse(RB090 == 2, "female", NA))) %>%
    filter(!is.na(sex)) %>%
    mutate(RB020 = ifelse(RB020 == "UK","GB",RB020),
           RB020 = ifelse(RB020 == "EL","GR",RB020),
           cntry = countrycode::countrycode(RB020, "iso2c", "country.name", warn = FALSE)) %>%
    select(RB010,RB020,RB030,PL031,RX010,max,age_cat,sex,cntry) %>%
    filter(sex == "female")#%>% head()

  library(dplyr)
  library(ggplot2)


  # Then we count we merge the domestics tasks+other inactivity, include only women and
  # count the lengt of periods for belonging to that class vs. not belonging.

  poverty_counts <- function(subset = "dat") {

    # Following loop processes one country at the time,
    # within one county, one year at time, and
    # a) calculates the poverty threshold
    # b) categories individual poverty status either TRUE or FALSE based on individual disposable household income

    df <- eval(parse(text = subset))
    # loop returns a data.frame with all individuals and whether they were poor or not at particular year
    datp <- data.frame()
    for (cntry in unique(as.character(df$RB020))){
      dd <- dat[dat$RB020 == cntry,]
      # for each year
      dp <- data.frame()
      for (year in unique(dd$RB010)) {
        dy <- dd[dd$RB010 == year,]
        #dy$th <- th
        dy$dom[dy$PL031 %in% 10:11] <- TRUE
        dy$dom[is.na(dy$dom)] <- FALSE
        dp <- rbind(dp,dy[c("RB010","RB020","RB030","dom")])
      }
      #assign(paste0("data_",cntry),dp)
      datp <- rbind(datp,dp)
    }

    # convert data from long to wide
    library(tidyr)
    datp <- datp[!duplicated(datp[c("RB010","RB020","RB030")]),]

    w <- spread(datp, # data
                RB010, # class-variable
                dom) # amount
    # add Y to variable names for easier processing in R
    # names(w) <- c("RB020","RB030","Y2007","Y2008","Y2009","Y2010","Y2011")
    names(w) <- c("RB020","RB030","Y2008","Y2009","Y2010","Y2011")
    # how many "missing data years" individuals have
    w$nas <- rowSums(is.na(w[3:6]))
    # if there is more than 1, individuals are dropped (we only include individuals with full four year record)
    w <- w[w$nas <= 1,]
    # how many TRUE years one has had?
    w$length <- rowSums(w[3:6],na.rm = TRUE)
    # print top of data
    #kable(head(w, 20))
    ##-----------------------------------

    # Country level proportions of different lenghts of poverty
    w$RB020 <- factor(w$RB020)
    tbl <- as.data.frame(prop.table(table(w$RB020,w$length),1)*100)
    # relabelling for plotting
    tbl$fil[tbl$Var2 == 0] <- "not poor"
    tbl$fil[tbl$Var2 == 1] <- "for 1 year"
    tbl$fil[tbl$Var2 == 2] <- "for 2 years"
    tbl$fil[tbl$Var2 == 3] <- "for 3 years"
    tbl$fil[tbl$Var2 == 4] <- "for 4 years"
    tbl$fil[tbl$Var2 == 5] <- "for 5 years"
    # reordering for plotting
    tbl$fil <- factor(tbl$fil,
                      levels=c("not poor",
                               "for 1 year",
                               "for 2 years",
                               "for 3 years",
                               "for 4 years",
                               "for 5 years"))

    # reordering countries for plotting based on non-poor share
    # tbl$cntry <- factor(tbl$cntry, levels=tbl[tbl$fil == "not poor",][order(tbl$Freq),]$cntry)
    list(tbl,w,df)
  }

  ls1 <- poverty_counts()

  library(dplyr)
  library(ggplot2)
  age <- dat %>% group_by(RB020,RB030) %>% summarise(max = max(RX010))

  w2 <- merge(ls1[[2]],age,by=c("RB020","RB030"), all.x=TRUE)
  w2$age_cat <- cut_interval(x = w2$max, length = 5)

  w2$RB020 <- factor(w2$RB020)
  dat1 <- data.frame()
  for (cntry in unique(as.character(w2$RB020))){
    d <- w2[w2$RB020 == cntry,]
    tbl1 <- as.data.frame(prop.table(table(d$age_cat,d$length),1)*100)
    tbl1$RB020 <- cntry
    dat1 <- rbind(dat1,tbl1)
  }

  # relabelling for plotting
  dat1$fil[dat1$Var2 == 0] <- "not poor"
  dat1$fil[dat1$Var2 == 1] <- "for 1 year"
  dat1$fil[dat1$Var2 == 2] <- "for 2 years"
  dat1$fil[dat1$Var2 == 3] <- "for 3 years"
  dat1$fil[dat1$Var2 == 4] <- "for 4 years"
  dat1$fil[dat1$Var2 == 5] <- "for 5 years"
  # reordering for plotting
  dat1$fil <- factor(dat1$fil,
                     levels=c("not poor",
                              "for 1 year",
                              "for 2 years",
                              "for 3 years",
                              "for 4 years",
                              "for 5 years"))

  dat1$cntry <- countrycode::countrycode(dat1$RB020, "iso2c", "country.name", warn = FALSE)

  # remove Iceland
  # dat1 <- dat1[dat1$cntry != "Iceland",]
  save(dat1, file="./local_data/eusilc_trimmed.Rdata")
} else load("./local_data/eusilc_trimmed.Rdata")

#%#% ---------------------------- #%#%##
#%#% manipulate data ------------ #%#%##
#%#% ---------------------------- #%#%##



dat1$Var1 <- str_replace_all(dat1$Var1, ",", " <~ " )
dat1$Var1 <- str_replace_all(dat1$Var1, "\\(", "")
dat1$Var1 <- str_replace_all(dat1$Var1, "\\)", "")
dat1$Var1 <- str_replace_all(dat1$Var1, "]", "")
dat1$Var1 <- str_replace_all(dat1$Var1, "\\[", "")

levels(dat1$fil)[levels(dat1$fil) == "not poor"] <- "not fulfilling \ndomestic tasks \nand care \nresponsibilities"

dat1$cntry <- as.character(dat1$cntry)

# dat1 <- dat1[!(dat1$cntry %in% "Iceland"),]
# dat1 <- dat1[!(dat1$cntry %in% "Norway"),]
dat1 <- dat1[!(dat1$cntry %in% "Luxembourg"),]

dat1$cntry <- factor(dat1$cntry, levels=c("Denmark","Finland","Sweden","Norway","Iceland",                               #
                                          "Austria","Belgium","France","Germany","Netherlands",   #
                                          "United Kingdom",                                                    #
                                          "Cyprus", "Greece","Italy", "Malta","Portugal","Spain",                                #
                                          "Bulgaria","Czech Republic","Estonia","Hungary","Latvia","Lithuania","Poland","Romania","Slovakia","Slovenia")
)

p <- ggplot(data=dat1,aes(x=Var1,y=Freq,fill=fil))
p <- p +  geom_bar(stat="identity", position = "stack")
# p <- p + scale_fill_manual(values=colorRampPalette(brewer.pal(9,"Oranges"))(6),
#                            guide = guide_legend(reverse=TRUE))
p <- p + scale_fill_grey(start =.85 , end = .1, guide = guide_legend(reverse=TRUE))
# p <- p + scale_fill_grey(start =.95 , end = .1, guide = guide_legend(reverse=TRUE))
p <- p + theme_minimal() + theme(legend.position = "right") +
  theme(text = element_text(family = fontti, size= 12)) +
  theme(legend.title = element_blank()) +
  theme(axis.text.y= element_text(size = 9)) +
  theme(axis.text.x= element_text(size = 7, angle=90, vjust= 0.5)) +
  theme(axis.title = element_text(size = 11, face = "bold")) +
  theme(legend.text= element_text(size = 11)) +
  theme(strip.text = element_text(size = 11, face = "bold")) +
  guides(colour = guide_legend(override.aes = list(size=4))) +
  theme(panel.border = element_rect(fill=NA,color="grey70", size=0.5,
                                    linetype="solid"))
p <- p + labs(x="", y= "Share of women (%)")
p <- p + facet_wrap(~cntry, ncol = 5)
```


```{r women, echo=FALSE, fig.width=9, fig.height=9, fig.pos="h!", fig.cap='Female labor-force status "taking care of family-related care responsibilities" (\\%); duration of care-related absences from the labor market according to age'}

print(p)

```

Figure 6 indicates the strong gender-based stratification effects in the labor markets and indicates that there still are substantial differences between the countries. A general trend is that female labor-force participation rates are getting higher and in more and more countries fathers have their specific paternity-leave schemes and daddy quotas leading to a more gender-neutral division of family responsibilities. This process has gone farthest in the Nordic countries, where mothers are liberated to work, and fathers are liberated to take on care responsibilities at home (Hobson, 2002; Ferrarini, 2006). However, this is apparently an unfinished revolution. As long as parental leave schemes are not designed as individual rights, it appears unlikely that the remaining gender differences will disappear not only from entry-exit patterns in the labor market but also from pay gaps and career opportunities. These gaps continue to exhibit gender deficits, which is contrary to the superior performance of women at all levels of the education systems in so many of the most advanced industrial nations. 
